Gusto: a toolkit for compatible finite element dynamical cores

### 2024-03-08 source/paper.tex
some edits and most of the pressure gradient section

\def\MM#1{\boldsymbol{#1}}
\newcommand{\pp}[2]{\frac{\partial #1}{\partial #2}} 
\newcommand{\JScomment}[1]{\textit{\textbf{Jemma says: #1}}}

@@

@@

\subsection{Compressible Euler Equations}
The dynamical core at the heart of any weather or climate model solves
the rotating compressible Euler equations that model atmospheric flow,
often with additional approximations such as shallow vs deep
atmosphere or hydrostatic vs non-hydrostatic flow (** mention this
here or at the end of the section??). We write these equations in
terms of the prognostic variables $\MM{u}$, $\rho$ and $\theta$ where
$\MM{u}=(u, v, w)$ is now a three dimensional velocity field, $\rho$
is the fluid density and $\theta$ is the potential temperature. In
terms of these variables, the equations are:

@@

\subsection{Finite Element Discretisation}
\label{sec: FEM}

@@

In line with the theory outlined in section \ref{}, we always choose
the velocity $\MM{u} \in \mathbb{V}_u$ with $\mathbb{V}_u \subset
H(\text{div})$. The field whose gradient appears in the `pressure'
gradient term is chosen to be in the space $\mathbb{V}_p$ with
$\mathbb{V}_p \subset L_2$. The compressible Boussinesq and
compressible Euler equations have a third prognostic variable, either
buoyancy or potential temperature, which is chosen to be in the space
$\mathbb{V}_\theta$ with $\mathbb{V}_\theta \subset ??$ These choices
are governed by the properties required of discretisation of the
linearised equation sets (as discussed in section ** and other
refs). To solve the nonlinear equation sets we need to introduce
stable and accurate discretisations of the transport terms and the
nonlinear pressure gradient term $c_p\theta\nabla \Pi$ in the
compressible Euler velocity equation. Since the prognostic variables
are discretised using different finite element function spaces
$\mathbb{V}_u$, $\mathbb{V}_p$ and $\mathbb{V}_\theta$, each with
different continuity properties between the elements, the choice of
stable transport discretisation is different for each prognostic
variable.

** Don't do whole equation set - go term by term and try to write as
few equations as possible.

@@

\JScomment{Text here depends on what the schematic looks like - this
  paragraph should refer to the schematic and explain all aspects} The
Gusto model comprises a domain and a prognostic equation set. The
domain holds the mesh and the finite element function spaces along
with any other information about the geometry such as the rotation
vector. The Prognostic equation sets are each defined in a python
class, making use of the common finite element forms for the transport
and difffusion terms. The timeloop class contains the definition of
the timestepping procedure and the \texttt{run} method that loops over
the timesteps. It also sets up diagnostic fields and calls the IO as
required. The simplest timeloop class will just apply one time
discretisation method to the equation set but it is common in weather
and climate prediction models to split the different types of terms
and apply a time discretisation separately to each type of term. For
example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
related to the physics are solved for in the appropriate part of the
iteration loop, with physics processes that occur on `fast' timescales
evaluated more frequently and `slower' physics processes evaluated
just once at the end of the timestep. The choice of where to evaluate
the physics processes in the iteration is also based on the
computational cost and the best way to do this is still an open
research question (**ref).

@@

To include:
\begin{itemize}
\item terminator toy (Tim)
\item shallow water on different grids and with different HDiv families (Nell and Dan)
\item convergence with different RK schemes / multistage schemes (Alex)
\item vertical slice Boussinesq / Euler (+ moisture?)
\item moist baroclinic wave in channel?? (Alex)
\item spherical gravity wave, Boussinesq (Jemma)
\item spherical baroclinic wave, Euler (Dan)
\end{itemize}

>>>>

\def\MM#1{\boldsymbol{#1}}
\DeclareMathOperator{\diff}{d}
\newcommand{\avg}[1]{\{#1\}}
\newcommand{\jump}[1]{[\![#1]\!]}
\newcommand{\pp}[2]{\frac{\partial #1}{\partial #2}} 
\newcommand{\JScomment}[1]{\textit{\textbf{Jemma says: #1}}}

@@

\begin{itemize}
\item Section 1: why compatible FEM
\item Section 2: why Gusto
\item Sections 3 and 4: this is Gusto
\item Section 5: and here are some examples of what we can currently do
\end{itemize}

@@

\subsection{Compressible Euler Equations}
The dynamical core at the heart of any weather or climate model solves
the rotating compressible Euler equations that model atmospheric
flow. We write these equations in terms of the prognostic variables
$\MM{u}$, $\rho$ and $\theta$ where $\MM{u}=(u, v, w)$ is now a three
dimensional velocity field, $\rho$ is the fluid density and $\theta$
is the potential temperature. The pressure gradient term is written in
terms of the Exner pressure $\Pi$ which is calculated through an
equation of state. In terms of these variables, the equations are:

@@

\section{Finite Element Discretisation}
\label{sec: FEM}

@@

In line with the theory outlined in section \ref{}, we always choose
the velocity $\MM{u} \in \mathbb{V}_u$ with $\mathbb{V}_u \subset
H(\text{div})$. The field whose gradient appears in the `pressure'
gradient term is chosen to be in the space $\mathbb{V}_p$ with
$\mathbb{V}_p \subset L_2$. The compressible Boussinesq and
compressible Euler equations have a third prognostic variable, either
buoyancy or potential temperature, which is chosen to be in the space
$\mathbb{V}_\theta$ which has degrees of freedom that are collocated
with those of the vertical velocity, mimicking the Charney-Phillips
staggering as discussed in **. These choices are governed by the
properties required of discretisation of the linearised equation sets
(as discussed in section ** and other refs). To solve the nonlinear
equation sets we need to introduce stable and accurate discretisations
of the transport terms and the nonlinear pressure gradient term in the
compressible Euler equation set. In addition, some test cases require
the addition of viscous and diffusive terms to generate converged
solutions \ref{}. In the following subsections we illustrate a
non-exhaustive (**probably!) selection of the options available in
Gusto.

\subsection{Transport terms}
Since the prognostic variables are discretised using different finite
element function spaces $\mathbb{V}_u$, $\mathbb{V}_p$ and
$\mathbb{V}_\theta$, each with different continuity properties between
the elements, the choice of stable transport discretisation is
different for each prognostic variable.

\subsection{Pressure gradient term}
The pressure gradient term is discretised in a similar way for each
equation set, with some additional considerations required for the
nonlinear form of this term in the compressible Euler equations. Since
the pressure (or depth, for shallow water) field is discretised in
$\mathbb{V}_p \subset L_2$, the gradient is not defined globally and
the term must be integrated by parts. In the linear case, using the
notation for the compressible Boussinesq equations, the weak form is

\begin{equation}
\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x,
\end{equation}
where $\Omega$ denotes the domain and there are no surface terms from
the integration by parts because these vanish on interior element
boundaries since the normal component of $\MM{w}$ is continuous across
element faces and on exterior boundaries we set no normal flow
boundary conditions. However, the pressure gradient term in the
compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
this introduces an additional complication. Integrating by parts is
still required since $\Pi$ is evaluated as a function of the density
$\rho \in L_2$ and so is discontinuous, but this means that we now
have a differential operator applied to $\theta \in \mathbb{V}_\theta$
which is discontinuous in the vertical. Following \ref{} we integrate
by parts in each element $e$, giving

\begin{equation}
  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\theta^-\MM{w}\cdot\MM{n}\Pi^- \diff S,
\end{equation}
where $\partial e$ denotes the element faces, $\MM{n}$ is the outward
pointing normal to the face and $\cdot^-$ denotes the value of a
variable on the interior side of the boundary. Summing over the domain
gives

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S,
\end{equation}
where $\jump{\cdot}$ and $\avg{\cdot}$.

\subsection{Viscous and diffusive terms}

** Don't do whole equation set - go term by term and try to write as
few equations as possible: transport, pressure gradient, diffusion

@@

\JScomment{Text here depends on what the schematic looks like - this
  paragraph should refer to the schematic and explain all aspects} The
Gusto model comprises a domain and a prognostic equation set. The
domain holds the mesh and the finite element function spaces along
with any other information about the space-time domain such as the
coordinates and the timestep. The Prognostic equation sets are each
defined in a python class, making use of the common finite element
forms for the transport and difffusion terms. The timeloop class
contains the definition of the timestepping procedure and the
\texttt{run} method that loops over the timesteps. It also sets up
diagnostic fields and calls the IO as required. The simplest timeloop
class will just apply one time discretisation method to the equation
set but it is common in weather and climate prediction models to split
the different types of terms and apply a time discretisation
separately to each type of term. For example, the transport and
forcing terms might be solved for separately, such as in **, which
enables explicit transport schemes to be applied to each field. **
something about SIQN structure. Terms related to the physics are
solved for in the appropriate part of the iteration loop, with physics
processes that occur on `fast' timescales evaluated more frequently
and `slower' physics processes evaluated just once at the end of the
timestep. The choice of where to evaluate the physics processes in the
iteration is also based on the computational cost and the best way to
do this is still an open research question (**ref).

@@

To include:
\begin{itemize}
\item terminator toy (Tim)
\item shallow water on different grids and with different HDiv
  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
  icosahedral sphere PV, high res (ref 6)
\item convergence with different RK schemes / multistage schemes using
  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
  one Linf) (Alex)
\item vertical slice Euler - Straka, different orders, illustrates
  recovery for diffusion (Tom)
\item 3D moist bubble with rain (Tom/Alex)
\item spherical gravity wave, Boussinesq (Jemma)
\item spherical baroclinic wave, Euler - whatever people usually plot
  (horizontal slice pressure contours, temperature shadng) (Dan)
\item Held Suarez??
\end{itemize}
